Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-04-17 main.tex
Still merging...

There are thus two important levels of concurrency that need to be exploited to
efficiently use a modern CPU: {\em core-level} concurrency and {\em vector-level}
concurrency.

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
on-chip cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
An explanation of the CUDA programming model at a useful level of completeness would take 
more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities, and refer the reader to the NVIDIA C for CUDA Programming Guide.

Our target GPU architecture
is the Nvidia Fermi architecture (\eg, the GTX 480 GPU used in this paper). 
CUDA based GPUs are comprised of several \emph{streaming multiprocessors} (SMP) 
(currently 14 to 16 for a high end GPU).  Each SMP is capable of executing up to 64
floating point operations concurrently.

%The GPU uses a SPMD (Single Program Multiple Data) programming model: at any given
%point in time, all threads are running the same gpu program, or \emph{kernel}, but
%they all operate on different data.  During kernel execution, threads can only communicate
%(or synchronize with each other) within groups of threads called blocks.  Each block
%is further subdivided into groups of 32 threads called \emph{warps}. 
%Typically, an SMP will hold several thread blocks at a time, although it "only" has enough 
%hardware to schedule
%2 warps to perform a floating point operation at a time (the remaining warps are either
%inactive, or waiting for data to arrive).  Threads within a block are able to synchronize
%(wait at certain points of a program), which gives them a chance to communicate via a
%relatively small amount of shared cache.

% GPU CACHE DISCUSSION
% for L2 cache
% 768KB = 786432 bytes
% 64*64*4 = 2**14 bytes = 16384 bytes per image
% 48 images
Each SMP contains its own L1-cache, which is divided between hardware managed
and software managed ("shared") memory.  Additionally, all SMPs share a common
L2-cache.  For our system, each SMP has 64 KB of L1 cache for a total of 896
KB of L1 cache/shared memory and 768KB of L2 cache. The relatively small amount
of cache (1/23 as much as CPU L3) on the GPU is balanced by a significantly
higher bandwidth between the processor chip and off-chip memory (DRAM) compared
to the CPU.  A scale drawing of the caches available on the GPU can be seen in Figure \ref{fig:caches}.

The GPU has its own memory system, any data the GPU uses must first be
transferred from CPU DRAM to GPU DRAM over PCI-Express.  For our application,
this transfer overhead is amortized over a large amount of computation and is
not a major concern.

\section{Parallelism in the Face Alignment Stage}
\label{sec:alignment}
There are essentially two levels of parallelism available for exploitation in
for iterative alignment in the face recognition application.  The primary goal
of this section is to introduce these two levels of parallelism.  Additionally,
in order for some important aspects of our implementations to be clear later,
we need to establish how the iterative alignment stage works in some
detail.

The inputs of the iterative alignment stage are the test image $f(x,y)$ stacked into an $M$-
dimensional vector $\f$, and the aligned and stacked training images $A_i$ for
each training user $i$. 
The algorithm repeatedly solves a linearized version of the minimization problem
in \eqref{eq:l1min_alignment}.  This involves two computational steps:
the computation of the linearization, and the solution of the minimization problem.

The computation of the linearization is complicated by the need to avoid trivial 
solutions where the algorithm
progressively zooms in on a black pixel of the test image.
This is achieved by minimizing the $\ell_1$ error between the illumination model $Ax$ and the cropped,
resampled, and normalized test image $\tilde\bb$, rather than the un-normalized version as in 
conventional Lucas-Kanade iterative alignment.
This involves a bit of calculus:
\begin{equation}
\frac{\partial}{\partial \tau_p} \frac{\bb}{||\bb||} 
= \frac{1}{||\bb||} \frac{\partial \bb}{\partial \tau_p} - 
\frac{\bb}{||\bb||^3} \bb^T \frac{\partial \bb}{\partial \tau_p}
\end{equation}
If the warping $\tau$ is constrained to be a similarity transform,
it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
once per test image.  The per-user iterative alignment algorithm is summarized as Algorithm \eqref{alg:iterative_alignment}.
\begin{algorithm}[thb]
\caption{\bf (Per-User Iterative Alignment)} \label{alg:iterative_alignment}
\begin{algorithmic}[1]
\begin{small}
\STATE {\bf Input:} Aligned training images $\{A_i \in \Re^{m\times n}\}$ for gallery subject $i$, the test image
$\f\in\Re^M$ 
%\STATE $\tau^{(0)} \leftarrow \textrm{face detector($\f$)}$. 
%\STATE $\f_x, \f_y \in \Re^M \leftarrow \frac{df}{dx}, \frac{df}{dy}$ 
%\STATE {\bf for} each gallery subject $i$, 
\STATE \hspace{0mm} {\bf while} not converged $(j=1,2,\ldots)$ {\bf do} 
\STATE \hspace{3mm} $\tilde \bb(\tau) \leftarrow \frac{\bb \circ \tau}{\|\bb \circ \tau\|_2}$;
\STATE \hspace{3mm} $J \leftarrow  \frac{\partial}{\partial \tau} \tilde\bb(\tau)  \bigr|_{\tau^{(j)}} $;
\STATE \hspace{3mm} $\Delta \tau =  \arg\min \; \| \e \|_1  \; \subj \; \tilde \bb + J \Delta \tau = A_i \x + \e.$ 
\STATE \hspace{3mm} $\tau^{(j+1)} \leftarrow \tau^{(j)} + \Delta
\tau$; 
\end{small}
\end{algorithmic}
\end{algorithm}

@@

We will show that, in constrast with previous implementations, which mapped
problem level parallelism onto core-level concurrency, it is more efficient to
completely disregard problem-level parallelism, and instead map pixel-level
parallelism onto both core-level and vector-level cpu concurrency.  In other
words, the cpu implementation we advocate performs alignment of the test image
against the gallery images of a single user at a time.  Each matrix-vector and
vector-vector operation gets broken into 8 parts that run one per core, and
into 4 single float vectors that execute on the SSE SIMD units.
This mapping of the alignment problem onto the concurrency provided by the pair
of CPUs is visualized in Figure \ref{fig:alignment_mapping_cpu}.
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_mapping_cpu}
\caption{Proposed mapping of alignment parallelism onto cpu concurrency}
\label{fig:alignment_mapping_cpu}
\end{figure}

@@

{\bf Implementation Details} In our implementation, most of the operations
required by the alignment stage take advantage of Intel Math Kernel Library
(MKL), a commercial implementation of the standard Basic Linear Algebra
Subprograms (BLAS).  For image resampling, we make use of Intel's Integrated
Performance Primitives (IPP) library.  Both of these libraries are heavily
optimized for Intel multicore CPU's, and are able to make good use of both
core-level and vector-level concurrency.  For operations that do not map well
onto existing libraries, we achieve both vector-level and core-level
concurrency through a combination of the OpenMP API \cite{dagum2002openmp} and
the automatic vectorization facilities of Intel's C++ compiler
(ICC)\cite{dulong1999overview}.

@@

This implementation was written using the OpenCV library
as a wrapper for MKL, and, like our proposed implementation, achieves vector-level concurrency implicitly.  
With eight alignment problems (associated with 8 training users) being solved
concurrently, and with 20-40 images per training user, all of the required data
fits into L3 cache.  While the L3 cache has much higher bandwidth and lower latency
than off-chip memory, this implementation is slower than the proposed implementation.

%During the course of development, many implementation strategies were tried.
%Listed in order of increasing performance based on a synthetic benchmark of
%the $\ell_1$ minimization step, they included:
%\begin{enumerate}
%\item A pure C++ implementation threaded and vectorized with OpenMP and icpc (200 ms).
%\item An OpenCV (with MKL backend) implementation threaded with OpenMP similar to \cite{WagnerA2011-PAMI} (190ms.)
%\item A straightforward MATLAB implementation (which uses threaded MKL internally) (60 ms.)
%\end{enumerate}

%\subsection{Tuning the GEMV operation}
%When matrix-vector operations are called sequentially on the same data,
%the operation can be blocked so that each thread operates on the same block of data for each operation.
%For Single threaded MKL sgemv, Code took          0.0335823 msec per instance to run!
%For Automatic multi-threaded MKL sgemv, Code took 0.008179 msec per instance to run!
%For Manually multi-threaded MKL sgemv, Code took  0.0061276 msec per instance to run!

%\subsection{Tuning the Soft-Thresholding Operator}
%...Make sure there are no branches
%...Make sure there are no function calls
%...Make sure the loop gets automatically vectorized
%...Block the computation into work for different cpu cores

@@

One possible implementation of the iterative alignment stage would be to
perform the optimization for all instances simultaneously, but to perform each
linear algebraic operation in a separate gpu kernel call.  The primary
advantage of this implementation strategy is that every kernel would be
performing a very simple operation requiring fewer registers per thread.  The
registers in a GPU are shared between all executing threads thus, barring other
hardware limitations, the fewer registers each thread requires, the more
threads can be scheduled to run simultaneously.  The primary downside of this
approach is that it forces synchronization between all problem instances.  It will
also have increased kernel launch overhead and potentially less efficient use of
the GPU's L2 cache.
% DREW Idea... in hindsight, maybe we should still try many kernel launches. 
% instances can still exit early, and can still update a global variable signaling
% that they're done

% Should really verify somehow that we're maxing out GPU bandwidth

>>>>

There are thus two important levels of concurrency that need to be exploited to
efficiently use a modern CPU: {\em core-level} concurrency and {\em sse-level}
concurrency.

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
on-chip cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
An explanation of the CUDA programming model at a useful level of completeness would take 
more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities\footnote{A discussion of the CUDA programming model can be found
in the NVIDIA C for CUDA Programming Guide}.

Our target GPU architecture is the Nvidia Fermi architecture (\eg, the GTX 480
GPU used in this paper).  CUDA based GPUs are comprised of several
\emph{streaming multiprocessors} (SMP), each of which is roughly analogous to a CPU core.
For Fermi architecture GPU's, there are up to 16 SMP's, and each SMP is capable of executing up to 64 single precision 
floating point operations
concurrently\footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}.

% GPU CACHE DISCUSSION
% for L2 cache
% 768KB = 786432 bytes
% 64*64*4 = 2**14 bytes = 16384 bytes per image
% 48 images
Each SMP contains its own L1-cache, which is divided between hardware managed
and software managed ("shared") memory.  Additionally, all SMPs share a common
L2-cache.  For our system, each SMP has 64 KB of L1 cache, and all SMP's share
768KB of L2 cache. 
%The cache hierarchy 
The relatively small amount
of cache (1/23 as much as CPU L3) on the GPU is balanced by a significantly
higher bandwidth between the processor chip and off-chip memory (DRAM) compared
to the CPU.  A scale drawing of the caches available on the GPU can be seen in Figure \ref{fig:caches}.
The GPU has its own memory system, and any data the GPU uses must first be
transferred from CPU DRAM to GPU DRAM over PCI-Express.  For our application,
this transfer overhead can be amortized over a large amount of computation and is
not a major concern.

While the programming model for
the SIMD units of each SMP is somewhat more flexible than on the CPU,
leveraging the flexibility typically comes at the cost of reduced parallelism
\footnote{In CUDA terms, code branches that cause warp divergence usually
result in serialization of the different groups of threads}.  Therefore, for
the purposes of comparing hardware architectures, the SIMD units on the GPU are
roughly analogous to the CPU SIMD units. Thus, in summary, the GPU hardware 
provides two levels of concurrency: {\em SMP-level}
concurrency and {\em warp-level} concurrency.  Note that the GPU provides more
concurrency than a cpu at both levels (14 SMP's vs. 8 cores) and (64 wide vs. 4 wide SIMD
units).

\section{Parallelism in the Face Alignment Stage}
\label{sec:alignment}

Given the hardware concurrency in modern parallel computing architectures, next,
we shall study the optimal mappings of the algorithm parallelism in face recognition to the multi-core CPUs and GPUs, respectively.
In this section, we first focus on the face alignment stage (see Figure \ref{fig:pipeline}).

Face alignment \eqref{eq:l1min_alignment} estimates an image transformation $\tau$ that rectifies the query image $\tilde{\bb}$ with possible pose variation
w.r.t. each training class $A_i$, which leads to the minimal sparsity in error $\ee$ after the alignment. Note that directly solving \eqref{eq:l1min_alignment} would require
simultaneously estimating $\xx$, $\ee$, and $\tau$ typically on a nonlinear manifold, which renders the problem extremely difficult to solve exactly with many local minima.

To mitigate the problem, if the system can obtain to a good initial guess of the transformation $\tau$ (\eg, provided by a good face detector),
the optimal solution for $\tau$ can be sought iteratively by a sequence of linearized approximations $j=1, 2, \cdots$ as follows:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
If the warping $\tau$ is constrained to be a similarity transform,
it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
once per test image.  The per-user iterative alignment algorithm is summarized as Algorithm \eqref{alg:iterative_alignment}.
\begin{algorithm}[thb]
\caption{\bf (Per-User Iterative Alignment)} \label{alg:iterative_alignment}
\begin{algorithmic}[1]
\begin{small}
\STATE {\bf Input:} Aligned training images $\{A_i \in \Re^{m\times n}\}$ for gallery subject $i$, the test image
$\f\in\Re^M$ 
%\STATE $\tau^{(0)} \leftarrow \textrm{face detector($\f$)}$. 
%\STATE $\f_x, \f_y \in \Re^M \leftarrow \frac{df}{dx}, \frac{df}{dy}$ 
%\STATE {\bf for} each gallery subject $i$, 
\STATE \hspace{0mm} {\bf while} not converged $(j=1,2,\ldots)$ {\bf do} 
\STATE \hspace{3mm} $\tilde \bb(\tau) \leftarrow \frac{\bb \circ \tau}{\|\bb \circ \tau\|_2}$;
\STATE \hspace{3mm} $J \leftarrow  \frac{\partial}{\partial \tau} \tilde\bb(\tau)  \bigr|_{\tau^{(j)}} $;
\STATE \hspace{3mm} $\Delta \tau =  \arg\min \; \| \e \|_1  \; \subj \; \tilde \bb + J \Delta \tau = A_i \x + \e.$ 
\STATE \hspace{3mm} $\tau^{(j+1)} \leftarrow \tau^{(j)} + \Delta
\tau$; 
\end{small}
\end{algorithmic}
\end{algorithm}

@@

We will show that, in constrast with previous implementations, which mapped
problem level parallelism onto core-level concurrency, it is more efficient to
completely disregard problem-level parallelism, and instead map pixel-level
parallelism onto both core-level and sse-level cpu concurrency.  In other
words, the cpu implementation we advocate performs alignment of the test image
against the gallery images of a single user at a time.  Each matrix-vector and
vector-vector operation gets broken into 8 parts that run one per core, and
into 4 single float vectors that execute on the SSE SIMD units.
This mapping of the alignment problem onto the concurrency provided by the pair
of CPUs is visualized in Figure \ref{fig:alignment_mapping_cpu}.
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_mapping_cpu}
\caption{Proposed mapping of alignment parallelism onto cpu concurrency}
\label{fig:alignment_mapping_cpu}
\end{figure}

@@

{\bf Implementation Details} In our implementation, most of the operations
required by the alignment stage take advantage of Intel Math Kernel Library
(MKL), a commercial implementation of the standard Basic Linear Algebra
Subprograms (BLAS).  For image resampling, we make use of Intel's Integrated
Performance Primitives (IPP) library.  Both of these libraries are heavily
optimized for Intel multicore CPU's, and are able to make good use of both
core-level and sse-level concurrency.  For operations that do not map well
onto existing libraries, we achieve both sse-level and core-level
concurrency through a combination of the OpenMP API \cite{dagum2002openmp} and
the automatic vectorization facilities of Intel's C++ compiler
(ICC)\cite{dulong1999overview}.

@@

This implementation was written using the OpenCV library
as a wrapper for MKL, and, like our proposed implementation, achieves sse-level concurrency implicitly.  
With eight alignment problems (associated with 8 training users) being solved
concurrently, and with 20-40 images per training user, all of the required data
fits into L3 cache.  While the L3 cache has much higher bandwidth and lower latency
than off-chip memory, this implementation is slower than the proposed implementation.

@@

One possible implementation of the iterative alignment stage would be to
perform the optimization for all instances simultaneously, but to perform each
linear algebraic operation in a separate gpu kernel call.  The primary
advantage of this implementation strategy is that every kernel would be
performing a very simple operation requiring fewer registers per thread.  The
registers in a GPU are shared between all executing threads thus, barring other
hardware limitations, the fewer registers each thread requires, the more
threads can be scheduled to run simultaneously.  The primary downside of this
approach is that it forces synchronization between all problem instances.  It will
also have increased kernel launch overhead and potentially less efficient use of
the GPU's L2 cache.

### 2011-04-17 main_allen.tex
Still merging...

While solving \eqref{eq:l1min_denoise} can achieve exceptional
recognition accuracy on public databases, its success also relies on the assumption
that the query image is well aligned with the training images. In
\cite{WagnerA2009-CVPR}, the framework of SBC was extended to iteratively align
a query image to one subject class. The face alignment algorithm in turn solves
the following minimization problem:
\begin{equation}
\hat{\tau}_i = \arg\min_{\xx, \ee, \tau_i} \|\ee\|_1\quad \mbox{subj. to}\quad \tilde\bb\circ\tau_i = A_i\xx + \ee,
\label{eq:l1min_alignment}
\end{equation}
where $\tau_i\in T$ is in a finite-dimensional group $T$ of transformations
(e.g., affine or homography) acting on the image domain, which aligns $\tilde\bb$
with the training (frontal) images from the $i$-th class $A_i$. The algorithm 
extends Lucas-Kanade iterative alignment \cite{LucasB1981} using the
$\ell_1$-norm as a robust error function.

@@

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a new face
recognition system has been fully implemented on a state-of-the-art
workstation for security and access control applications, as shown in Figure \ref{fig:pipeline}. 
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:parallelism}.
Finally, note that for applications targeting thousands of users or more, it would likely be
necessary to distribute the work over a network of computers, while this paper
focuses on the (shared memory) parallelism available in a single machine.
The C++ source code of our implementations, which target Nvidia's CUDA GPU architecture,
will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

@@

Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves 
computationally expensive operations such as matrix-matrix multiplication and 
the solution of least-squares problems with varying $A$ matrices. 
In Section \ref{sec:ALM}, we contend
that ALM is a better choice among existing $\ell_1$-min
solvers for implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) methods \cite{WrightS2008,BeckA2009}.
IST algorithms mainly involve elementary operations such as vector 
algebra and matrix-vector multiplication. Therefore,
when the dimension of the problem becomes high, IST-type algorithms are
particularly suitable for systems with parallel-computing pipelines. In
\cite{YangA2010-ICIP}, the authors further showed that, compared to other
IST-type algorithms, ALM is able to achieve good estimation accuracy comparable
to BP and Homotopy, while at the same time significantly improve the speed
compared to these exact $\ell_1$-min solutions. Due to these facts, in this
paper, we choose ALM as the core algorithm to implement the complete parallel
face recognition pipeline.

In terms of the past works in parallel $\ell_1$-min, the literature has been
limited, to the best of our knowledge. In \cite{BorghiA2010}, Borghi et al.
developed a special proximal gradient $\ell_1$-min algorithm based on
Moreau-Yosida regularization. In \cite{MurphyM2010}, Murphy et al. presented
parallel implementation of the l1-SPIRIT MRI reconstruction algorithm on NVidia
programmable GPUs, which is similar to the hardware setup of this paper.

@@

\section{Hardware Concurrency} \label{sec:parallelism}

In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since on-chip
memory (cache) is often orders of magnitude faster than off-chip memory, the
size and speed of on-chip memory (cache) have big performance implications.
We give a brief overview of the caches that are available in our target
architectures, and defer discussion of their performance implications to
sections \ref{sec:alignment} and \ref{sec:recognition}.

We will primarily compare CPU and GPU architectures on a per-board basis:
for CPU implementation we make use of all the cores in as many CPUs are present
on one motherboard, and for GPU implementation we make use of a single
many-core Nvidia GPU chip on a PCI card. GPUs are commonly shipped with a
single GPU chip per PCI-express card.  Note that most engineering  workstations
are capable of supporting several GPU cards.  Unless otherwise specified, all
implementations utilize single precision floating point datatypes.

\subsection{CPU Hardware Concurrency} 
\label{sec:CPU-concurrency}
The main defining characteristics of contemporary multi-core CPU architectures
are that they have two levels of concurrency, relatively large amounts of
on-chip cache, and relatively high clock speed.

The baseline architecture for our experiment is a Linux workstation with two
quad-core Intel Nehalem E5530 processors clocked at 2.4 GHz.  Each processor
has its own memory interface, and is directly interfaced to half of the RAM
installed in the machine.  The amount of RAM installed exceeds the amount used
by the algorithms, and is not an important performance consideration.

% CPU CACHE DISCUSSION
%2MB = 2**21 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**7  = 128 images.
Each core has a private 32\,KB L1 data cache and a private 256\,KB L2 cache (\ie, a total of 2\,MB L2 cache available for the eight cores).
Therefore, for the low-resolution images used in face recognition (say, on the order of $64\times64$ grayscale images), the L2 cache
in principle can hold up to 16 images per core and 128 images in total.
%16MB = 2**24 bytes = 16777216 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**10  = 1024 images.
Each processor further has 8\,MB of L3 cache that is shared by the four cores.
Overall, the algorithm has approximately 16\,MB of L3 cache
available for a dual-processor configuration, which corresponds to about 1024 images.

%With Hyperthreading enabled, each core can support two threads simultaneously.
For floating-point instructions, each core also has a vector processing unit
(SSE) capable of performing the same arithmetic operation on four single-precision floating point values simultaneously.

There are thus two important levels of concurrency that need to be exploited to
efficiently use a modern CPU: {\em core-level} concurrency and {\em vector-level}
concurrency.

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
on-chip cache, and relatively low clock speed.

Compared to CPU architectures, modern GPU architectures dedicate less of their transistors
to cache, and more of them for arithmetic logic. Our target GPU architecture
is the Nvidia Fermi architecture (\eg, GTX 480 GPU used in this paper). CUDA based GPUs are comprised of several 
\emph{streaming multiprocessors} (SMP) (currently 14 to 16 for a high end GPU).

The GPU executes programs as SIMT (Single Instruction Multiple Threads) in
groups of 32 threads called \emph{warps}.  These warps are then grouped into
higher levels called \emph{thread blocks} and \emph{grids}.  Since SMPs are not designed
to communicate between themselves during GPU program execution, tasks
are typically broken up into multiple GPU programs (kernels) that are executed
serially.\footnote{Please add a footnote to further explain.}

Each SMP contains its own L1 cache, which is divided between hardware managed
and software managed (``shared'') memory.  Additionally, all SMPs share a common
L2 cache.  For our system, each SMP has 64 KB of L1 cache with a total of 896
KB of L1 cache/shared memory and 768KB of L2 cache. The relatively small amount
of cache (1/23 as much as CPU L3) on the GPU is ballanced by a significantly
higher bandwidth between the processor chip and off-chip memory (DRAM) compared
to the CPU.

Compared to typical CPUs, GPUs have much higher peak computational bandwidth
as well as data bandwidth to DRAM.  However, since our system primarily consists
of matrix-vector operations and image resampling operations, it has a very low
arithmetic intensity (\ie, we perform a very small number of arithmetic
operations for every number that has to be loaded), and thus performance is
typically constrained by memory (and cache) bandwidth.  At large problem sizes
the CPU is limited by its relatively low bandwidth to DRAM compared to the GPU,
while at small problem sizes the GPU is limited by the relatively high latency
of its access to DRAM.

>>>>

While solving \eqref{eq:l1min_denoise} can achieve exceptional
recognition accuracy on public databases, its success also relies on the assumption
that the query image is well aligned with the training images. In
\cite{WagnerA2009-CVPR}, the framework of SBC was extended to iteratively align
a query image to each subject class individually. The face alignment algorithm solves
the following minimization problem:
\begin{equation}
\hat{\tau}_i = \arg\min_{\xx, \ee, \tau_i} \|\ee\|_1\quad \mbox{subj. to}\quad \tilde\bb\circ\tau_i = A_i\xx + \ee,
\label{eq:l1min_alignment}
\end{equation}
where $\tau_i\in T$ is in a finite-dimensional group $T$ of transformations
(e.g., affine or homography) acting on the image domain.  The sequence
$\tau_i$ converges to a transformation that aligns the test image $\tilde\bb$
with the training images from the $i$-th class $A_i$. In other words, the algorithm 
extends Lucas-Kanade iterative alignment \cite{LucasB1981} to use the
$\ell_1$-norm as a robust error function, while simultaneously estimating
the coefficients of the linear illumination model $A\x$.

@@

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a new face
recognition system for security and access control applications has been fully implemented on a state-of-the-art
workstation, as shown in Figure \ref{fig:pipeline}. 
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that for applications targeting thousands of users or more, it would likely be
necessary to distribute the work over a network of computers, while this paper
focuses on the parallelism available in a single (shared memory) machine.
The C++ source code of our implementations, which target Nvidia's CUDA GPU architecture,
will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

@@

Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves 
computationally expensive operations such as matrix-matrix multiplication and 
the solution of least-squares problems with varying $A$ matrices. 
In Section \ref{sec:ALM}, we contend
that ALM is a better choice for implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) methods \cite{WrightS2008,BeckA2009}.
IST algorithms mainly involve elementary operations such as vector 
algebra and matrix-vector multiplication. Therefore,
when the dimension of the problem becomes high, IST-type algorithms are
particularly suitable for hardware systems with a high degree of concurrency. In
\cite{YangA2010-ICIP}, the authors further showed that, compared to other
IST-type algorithms, ALM is able to achieve good estimation accuracy comparable
to BP and Homotopy, while at the same time significantly improve the speed
compared to these exact $\ell_1$-min solutions. Due to these facts, in this
paper, we choose ALM as the core algorithm to implement the complete parallel
face recognition pipeline.

In terms of the past works in parallel $\ell_1$-min, the literature has been
limited, to the best of our knowledge. In \cite{BorghiA2010}, Borghi et al.
developed a special proximal gradient $\ell_1$-min algorithm based on
Moreau-Yosida regularization. In \cite{MurphyM2010}, Murphy et al. presented
parallel implementation of the l1-SPIRIT MRI reconstruction algorithm on NVidia
CUDA programmable GPU architecture, which is also used in this paper.

@@

\section{Hardware Concurrency} \label{sec:concurrency}

In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since on-chip
memory (cache) is often orders of magnitude faster than off-chip memory, the
sizes and speeds of on-chip memory (cache) have big performance implications.
We give a brief overview of the caches that are available in our target
architectures, and defer discussion of their performance implications to
sections \ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

We will primarily be comparing CPU and GPU architectures on a per-board basis:
for CPU implementation we make use of all the cores in as many CPUs are present
on one motherboard, and for GPU implementation we make use of a single
many-core Nvidia GPU chip on a PCI card. GPUs are commonly shipped with a
single GPU chip per PCI-express card.  Note that most engineering  workstations
are capable of supporting several GPU cards.  Unless otherwise specified, all
implementations utilize single precision floating point datatypes.

\subsection{CPU Hardware Concurrency} 
\label{sec:CPU-concurrency}
The main defining characteristics of contemporary multi-core CPU architectures
are that they have two levels of concurrency, relatively large amounts of
on-chip cache, and relatively high clock speeds.

The baseline architecture for our experiments is a Linux workstation with two
quad-core Intel Nehalem E5530 processors clocked at 2.4 GHz.  Each processor
has its own memory interface, and is directly interfaced to half of the RAM
installed in the machine.  The amount of RAM installed exceeds the amount used
by the algorithms, and is not an important performance consideration.

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figures/caches}
\end{center}
\caption{Scale Drawing of the CPU and GPU cache sizes}
\label{fig:caches}
\end{figure}

% CPU CACHE DISCUSSION
%2MB = 2**21 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**7  = 128 images.
Each core has a private 32\,KB L1 data cache and a private 256\,KB L2 cache.
%16MB = 2**24 bytes = 16777216 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**10  = 1024 images.
Each processor further has 8\,MB of L3 cache that is shared by the four cores.
Overall, the algorithm has approximately 16\,MB of L3 cache
available for a dual-processor configuration.

For floating-point instructions, each core also has a vector processing unit
(SSE) capable of performing the same arithmetic operation on four single-precision 
floating point values simultaneously \footnote{By the publication date, the vector
units of typical high-end CPUS will double.}.

There are thus two important levels of concurrency that need to be exploited to
efficiently use a modern CPU: {\em core-level} concurrency and {\em sse-level}
concurrency.

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
on-chip cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
An explanation of the CUDA programming model at a useful level of completeness would take 
more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities\footnote{A discussion of the CUDA programming model can be found
in the NVIDIA C for CUDA Programming Guide}.

Our target GPU architecture is the Nvidia Fermi architecture (\eg, the GTX 480
GPU used in this paper).  CUDA based GPUs are comprised of several
\emph{streaming multiprocessors} (SMP), each of which is roughly analogous to a CPU core.
For Fermi architecture GPU's, there are up to 16 SMP's, and each SMP is capable of executing up to 64 single precision 
floating point operations
concurrently\footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}.

% GPU CACHE DISCUSSION
% for L2 cache
% 768KB = 786432 bytes
% 64*64*4 = 2**14 bytes = 16384 bytes per image
% 48 images
Each SMP contains its own L1-cache, which is divided between hardware managed
and software managed ("shared") memory.  Additionally, all SMPs share a common
L2-cache.  For our system, each SMP has 64 KB of L1 cache, and all SMP's share
768KB of L2 cache. 
%The cache hierarchy 
The relatively small amount
of cache (1/23 as much as CPU L3) on the GPU is balanced by a significantly
higher bandwidth between the processor chip and off-chip memory (DRAM) compared
to the CPU.  A scale drawing of the caches available on the GPU can be seen in Figure \ref{fig:caches}.
The GPU has its own memory system, and any data the GPU uses must first be
transferred from CPU DRAM to GPU DRAM over PCI-Express.  For our application,
this transfer overhead can be amortized over a large amount of computation and is
not a major concern.

While the programming model for
the SIMD units of each SMP is somewhat more flexible than on the CPU,
leveraging the flexibility typically comes at the cost of reduced parallelism
\footnote{In CUDA terms, code branches that cause warp divergence usually
result in serialization of the different groups of threads}.  Therefore, for
the purposes of comparing hardware architectures, the SIMD units on the GPU are
roughly analogous to the CPU SIMD units. Thus, in summary, the GPU hardware 
provides two levels of concurrency: {\em SMP-level}
concurrency and {\em warp-level} concurrency.  Note that the GPU provides more
concurrency than a cpu at both levels (14 SMP's vs. 8 cores) and (64 wide vs. 4 wide SIMD
units).
